---
title: "ST2195 Coursework Part 1 Code (Rmd), ID: 240626323"
author: "StevenSio"
date: "2024-11-16"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Import libraries

```{r}
library(ggplot2)
```

Probability Density Function

```{r}
f <- function(x) {
    # define Probability density function
    return(0.5 * exp(-abs(x)))
}
```

Function for Random Walk Metropolis Algorithm

```{r}
rwm_sample <- function(x, s, N) {
    # function to generate N random variables using Metroplis-Hastings Algorithm
    # x = initial value for x
    # s = standard deviation
    # N = number of trials

    # Our initial value for the algorithm
    previous_number <- x
    sample <- c() # empty vector to store values
    for (i in 1:N) {
        # Draw random number in a normal distribution with previous number as mean and sd of s
        new_number <- rnorm(1, previous_number, s)
        # compute log ratio of PDFs
        log_ratio <- log(f(new_number)) - log(f(previous_number))
        # draw random number in uniform distribution
        u <- runif(1, 0, 1)
        # if log u is lower than log ratio, keep the new number in the sample, else use the previous number
        if (log(u) < log_ratio) {
            sample[i] <- new_number
        } else {
            sample[i] <- previous_number
        }
        # set current number as the new previous number of the next iteration
        previous_number <- sample[i]
    }
    return(sample)
}
```

Generate 10,000 samples, with s=1 and initial value x=1.

```{r}
set.seed(1)

sample <- rwm_sample(0, 1, 10000)

# Look at the first 10 instances
print(sample[1:10])

# check the length
print(length(sample))
```

### Plotting the sample:

I used ggplot's `after_stat` parameter to plot the kernel density after transforming the data.

```{r}
# Sample mean and standard deviation
sample_mean <- mean(sample)
sample_sd <- sd(sample)

# Create a data frame for ggplot
sample_df <- data.frame(value = sample)

# Combined histogram and density plot
ggplot(data = sample_df, aes(x = value)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30, fill = "lightgray", color = "black", alpha = 0.6) + 
  geom_density(color = "blue") +
  stat_function(fun = f, color = "red") +
  labs(
    y = "Density",
    title = paste("Monte Carlo estimate of f(x)\nSample Mean:",
                  round(sample_mean, 3), "Standard Deviation:",
                  round(sample_sd, 3))
  )

```
### Obtain Convergence Diagnostic.

First I define a function to generate a matrix of J chains using Random Walk Metropolis algorithm, with J initial values for the algorithm.
```{r}
chain_samples <- function(x, s, N) {
    # function to generate a matrix storing J amount of chains using Random Walk Metropolis algorithm.
    # x = a list of initial values for each chain generation process.
    # s = standard deviation of RWM number generating algorithm.
    # N = size of each chain.

    # empty vector which is used to store all values from the samples
    chains <- c()
    for (i in 1:length(x)) {
        # store the generated vector in a temporary vector first
        temp_vector <- rwm_sample(x[i], s, N)
        # append the temp vector into the chains vector
        chains <- append(chains, temp_vector)
    }
    # format the whole chains vector into a matrix, with J amount of columns, each column represents a chain.
    chains <- matrix(chains, byrow = FALSE, ncol = length(x))
    return(chains)
}
```

Using the function to generate a chain of 4 samples with: 

-  4 random initial values between 0 and 1
-  s = 0.001 - N = 2000

```{r}
set.seed(1)
x1 <- runif(1)
x2 <- runif(1)
x3 <- runif(1)
x4 <- runif(1)

chains <- chain_samples(c(x1,x2,x3,x4), 
                        s=0.001,
                        N=2000)

# look at first 10 values of the chains
chains[1:10,]
```
Function to compute each chain's within sample mean and variance and store it in a matrix.
```{r}
within_sample <- function(chains) {
    # function to compute within sample mean and variance and store it in a matrix.

    stats_vector <- c()
    # for each sample in the chain, compute its variance, append into a vector
    for (i in 1:dim(chains)[2])# 2nd dimension of matrix (ncol).
      { 
        mean <- mean(chains[, i])
        variance <- var(chains[, i])
        # store the stats in a temp
        temp_vector <- c(mean, variance)
        # append into the stats_vector
        stats_vector <- append(stats_vector, temp_vector)
    }
    # format the whole vector into a matrix, with J amount of columns, each column represents a chain.
    within_sample <- matrix(stats_vector, byrow = FALSE, ncol = dim(chains)[2])
    rownames(within_sample) <- c("mean", "variance")
    return(within_sample)
}

```

Checking values
```{r}
mean(matrix(chains[,1:4],byrow = TRUE))

print(within_sample(chains))
mean(chains[, 1])
var(chains[, 1])
```
### Function to compute R value (convergence diagnostic)
```{r}
R_value <- function(chains) {
    # function to compute R value for one chain
  
    # using the function defined earlier to get the within sample stats
    within_stats <- within_sample(chains)

    ## Overall within sample variance
    # compute sum of within sample variance
    sum_variance <- sum(within_stats[2,])
    # overall within sample variance = sum of within sample variance / number of samples
    overall_w_var <- sum_variance / dim(chains)[2]

    ## Overall within sample mean
    # compute sum of within sample mean
    sum_mean <- sum(within_stats[1,])
    # overall within sample mean = sum of within sample mean / number of samples
    overall_w_mean <- sum_mean / dim(chains)[2]

    ## Between sample variance
    # compute sum of (individual sample mean - overall within sample mean)^2
    between_sumsquared <- 0
    for (i in 1:dim(chains)[2]) {
        temp_sumsquared <- (within_stats[1, i] - overall_w_mean)^2
        # sum the sum of squares one by one
        between_sumsquared <- between_sumsquared + temp_sumsquared
    }
    # sum of squared variance / number of samples
    between_variance <- between_sumsquared / dim(chains)[2]
    #print(between_variance)
    
    ## R Value
    # sqrt(between var + overall within var/ overall within var)
    R_value <- sqrt((between_variance + overall_w_var) / overall_w_var)
    # remove the original name of the vector
    #names(R_value) <- NULL
    
    #print(paste('between',between_variance,'overall_w',overall_w_var))
    return(as.numeric(R_value))
}
```

R value for a chain of 4 samples: <br> with 4 random initial values, s = 0.001, N = 2000.

```{r}
set.seed(1)
chains1 <- chain_samples(x = c(x1,x2,x3,x4), s = 0.001, N = 2000)

print(R_value(chains1))
```

Keeping N and J fixed, plot values of R with s values in the interval between 0.001 and 1.
```{r}
set.seed(10)
x1 <- runif(1)
x2 <- runif(1)
x3 <- runif(1)
x4 <- runif(1)
# Generate 1000 equally spaced values from 0.001 to 1
s <- seq(0.001, 1, length.out = 1000)

# print initial values
print(paste(x1,x2,x3,x4))

# Initialize a empty vector to store the R Values
R_values <- c()

for (sd in s) {
    chains <- chain_samples(x = c(x1,x2,x3,x4), s = sd, N = 2000)
    # temp variable
    temp_R <- R_value(chains)
    # append temp into R_values vector
    R_values <- append(R_values, temp_R)
}
# print first 10 R values
print(R_values[1:10])
```

Plot
```{r}
R_value_df <- data.frame(R_value = R_values, s = s)

ggplot(R_value_df, aes(x = s, y = R_value)) +
    geom_line() +
    geom_hline(aes(yintercept = 1.05), linetype = "dashed")+
    labs(title = "R value for different value of s")
```
